Exploring Constraint Grammar as a SAT problem

### 2015-03-26 paper/nodalida2015.tex
more text; now at 831 words

@@

We represent Constraint Grammar \cite{karlsson1995constraint} 
as a Boolean satisfiability (SAT) problem.
%This increases the computational complexity, 
%but we are exploring ways in which it could simplify the grammar writing:
%our experiments show that less rules are needed, and features like ordering of the rules
%and requirement of being unambiguously tagged become irrelevant.
% Our implementation is inspired by \cite{lager98,lager_nivre01},
%but we try to emulate the CG formalism more closely than the previous logic-based works.
Let us demonstrate our approach with the following example in Spanish.

@@

\begin{itemize}
\item The condition of being unambiguously tagged becomes irrelevant.
Rather than waiting for a word to get disambiguated, the SAT solver starts by 
making assumptions (e.g. ``\emph{casa} is a noun'') and working under them,
discarding the assumption if it doesn't lead to a model that satisfies all constraints.

\item Rules are unordered. 
\item Rules interact with each other and disambiguate more than same rules in the standard implementation.
\end{itemize}

%\section{CG as a SAT problem}
%\label{sect:satcg}

%SAT solving is based on a technique called unit propagation:
%a set of complex Boolean clauses is simplified, starting from unit
%clauses, which consist of just a single variable, and working up to a
%solution, where all variables in the set have a True or False value.
%We model an ambiguous sentence as a set of clauses, where each analysis is a variable, 
%and constraint rules form clauses operating on those variables.

\section{Rule ordering}
\label{sec:ordering}

A typical CG rule sequence starts with weaker rules, and leaves stronger rules towards the end, 
to be applied if none of the weaker rules has fired.

Unordered rules are not new: finite-state implementations of CG \cite{koskenniemi92} also have logically unordered rules.

% VISLCG ad hoc ordering:
% 1. SELECT before REMOVE.
% 2. SELECT rules targetting more preferred tags before 
%   rules targetting less preferred targets.
% 3. REMOVE rules targetting less preferred tags before
 %   rules targetting more preferred targets.
% 4. By order of appearance in rule file.

In CG3, the rules are applied in the order of appearance in the rule file.

Different strategies for maximising: ...

\paragraph{Rules can be flipped.}
Since implications can be presented as disjunctions, we can 
The representation of the implications as disjunctions 
\begin{itemize}
\item [] \texttt{REMOVE v IF -1 det} \\ $\rightarrow$ \texttt{REMOVE det IF 1 verb}   
\item [] \texttt{REMOVE adv IF NOT 1C adj} \\ $\rightarrow$ \texttt{REMOVE (*) - Adj  IF -1 adv}  \\
\end{itemize}

With this experiment, we move even further away from a direct ``if this then that'' model of CG.

@@

Application of logic to POS tagging or shallow parsing isn't new;
\cite{lager98,lager_nivre01} present constraint rules as 
a disjunctive logic program.
\cite{lindberg_eineborg98ilp,asfrent14} use Inductive Logic Programming to learn CG rules from a tagged corpus.
\cite{lager01transformation} discusses ordering of the CG rules.

@@

SAT is NP-complete, whereas the current implementations of CG are polynomial, 
but with advances in SAT solving techniques, the performance isn't as catastrophically bad as in 90s-00s.
We used the open-source SAT solver MiniSat \cite{een04sat}.
We do foreach sentence: foreach rule; 
%whereas VISL CG3 has foreach rule: foreach sentence.
the size of the SAT problem is dependent of the sentence (???).
TODO: time it comparing it to VISL CG3.
Note also that this is a first proof-of-concept implementation and could be optimised.

@@

\begin{table}
  \centering
  \begin{tabular}{|l|c|c|}
     \hline
    \textbf{Text} & \textbf{Words} & \textbf{\% same analyses} \\
    \hline
	Pride and Prejudice & 153386 & 90.6 \\  
    \hline
       Don Quijote & 186460 & 96.3  \\ 
     \hline
  \end{tabular}
  \caption{Results for comparing SATCG and VISLCG3.}
  \label{table:results}
\end{table}

The tendency of needing less rules to disambiguate shows also in a larger context with more complex rules.
SATCG disambiguated more 17 % of the cases for P&P,

>>>>

\usepackage{color}
\newcommand{\todo}[1]{{\color{cyan}\textbf{[TODO: }#1\textbf{]}}}

@@

We represent Constraint Grammar \cite{karlsson1995constraint} 
as a Boolean satisfiability (SAT) problem.
%This increases the computational complexity, 
%but we are exploring ways in which it could simplify the grammar writing:
%our experiments show that less rules are needed, and features like ordering of the rules
%and requirement of being unambiguously tagged become irrelevant.
Let us demonstrate our approach with the following example in Spanish.

@@

\paragraph{Condition of being unambiguously tagged becomes irrelevant.}
Rather than waiting for a word to get disambiguated, the SAT solver starts by 
making assumptions (e.g. ``\emph{casa} is a noun'') and working under them,
discarding the assumption if it doesn't lead to a model that satisfies all constraints.

\paragraph{Rules disambiguate more.}
% Rules interact with each other and disambiguate more than same rules in the standard implementation. 
In standard CG, the rules only tell to remove readings for the word which is tagged as 
\texttt{prn} or \texttt{det}. The SAT-based implementation interprets the rules as ``noun and pronoun together are illegal'', and is free to take action that concerns also the following word.
This means that the rules can be flipped: compare the rules below to the ones in the example.

\begin{itemize}
\item [] \texttt{REMOVE n IF (-1 prn) ;} \\
             \texttt{REMOVE vblex IF (-1 det) ;}
\end{itemize}

SAT-based approach gives identical results with both sets of rules,
whereas the standard CG would remove one reading from \emph{casa} and leave \emph{la} ambiguous.

\todo{Testing this property with more text, more complex rules and a larger rule set. }

\paragraph{Rules are unordered.}
A typical CG rule sequence starts with rules that are either very common, or have a complex condition to fulfil.
Stronger rules appear towards the end, to be applied if none of the weaker rules has fired. 
For a rule set such as following, this order is crucial:

\begin{itemize}
\item [] \texttt{SELECT Year IF (-1 Seasons OR Months) ;} \\
             \texttt{REMOVE Year ;}
\end{itemize}

Finite-state implementations of CG \cite{koskenniemi92} \footnote{\todo{Inari: check Tapanainen 1996 if there's anything about ordering}} also have logically unordered rules.

To make up for the lack of sequential order, we have experimented with different strategies of maximising and solving conflicts:
\begin{itemize}
\item Number of rules used
\item Number of instances when a rule is applied
\item 
\end{itemize}

@@

Our work is inspired by \cite{lager98}, which presents constraint rules as a disjunctive logic program,
and \cite{lager_nivre01}, which reconstructs 4 different formalisms from a logical point of view.
\cite{lindberg_eineborg98ilp,asfrent14} use Inductive Logic Programming to learn CG rules from a tagged corpus.
\cite{lager01transformation} discusses ordering of the CG rules.

@@

SAT is NP-complete, whereas the current implementations of CG are polynomial, 
but with advances in SAT solving techniques, the performance isn't as catastrophically bad as in 90s-00s.
We used the open-source SAT solver MiniSat \cite{een04sat}.
We do foreach sentence: foreach rule; 
%whereas VISL CG3 has foreach rule: foreach sentence.
the size of the SAT problem is dependent of the sentence (???).

The table below shows
Note also that this is a first proof-of-concept implementation and could be optimised.

\begin{table}
  \centering
  \begin{tabular}{|l|c|c|}
     \hline
	         & Quijote      & Pride \\ \hline
  VISL CG3 & 0m10.745s & 0m4.811s \\ \hline
  SATCG    & 2m53.708s & 0m55.705s \\ \hline
  \end{tabular}
  \caption{Comparing time between SATCG and VISL CG3.}
  \label{table:time}
\end{table}

@@

\begin{table}
  \centering
  \begin{tabular}{|l|c|c|}
     \hline
    \textbf{Text} & \textbf{Words} & \textbf{\% same analyses} \\
    \hline
	Pride and Prejudice & 153386 & 90.6 \\  
    \hline
       Don Quijote & 186460 & 96.3  \\ 
     \hline
  \end{tabular}
  \caption{Comparing analyses between SATCG and VISL CG3.}
  \label{table:results}
\end{table}

The tendency of needing less rules to disambiguate shows also in a larger context with more complex rules.
SATCG disambiguated more 17 \% of the cases for Pride and Prejudice, and 28 \% of the cases for Don Quijote.
